Is high social class always beneficial for survival? A study of adults and

### 2017-09-09 article2.Rmd
so what?

A major theme in demographic and epidemiological studies, is the seemingly persistent effect of social class on mortality. In the present study, we challenge this notion by taking a very long-term perspective on the development of social inequalities in mortality in the adult and elderly population with a special focus on the relative differences. The arguments for our statements is based on an investigation of the Skellefteå and Umeå regions in the north of Sweden for the period 1801--2013 for Skellefteå and 1901--2013 for Umeå. The main issue is then analysed according to gender and age-group (working age vs retired) and major causes of death (cardiovascular diseases and cancers). Furthermore we put this in the context of how the inequality in mortality is associated with the development of economic inequality in society. The results are discussed in relation to our understanding of the mortality transition and the social determinants of health and mortality, as well as their implications on some of the most influential hypotheses and concepts in health research.

On the basis of the results we present, we argue that high social class is not necessarily always favourable for survival. Mortality risks in different contexts must be understood in the intersection between class and gender. We suggest that health-related behaviour was important not only in present-day societies, but were decisive also in earlier phases of the mortality transition. The results implicate that the association between social class and health is more complex than is assumed in many of the dominant theories in demography and epidemiology.

## Social class and mortality

@@

Another practical problem is the size of that data set:
It contains no less than `r as.integer(round(n.exp + pn.exp))` person years and
 `r n.d + pn.d` deaths.
This makes it very inconvenient to use the nowadays so popular semi-parametric
technique known as
\emph{Cox regression} [@cox72, @gb02]. Instead we use a parametric model
that allows for reduction of the data
to \emph{sufficient statistics} [@fisher22]. The idea is to keep all
relevant covariates \emph{categorical}, and then create a table with all
possible combinations of levels of the covariates. For each such
combination, the total amount of exposure and the number of deaths are
calculated.  We have two sexes, three marital statuses, ten periods, two
urban/rural groups, four social classes at age 40,  four social classes at age 60,
ten age groups, and 22 cohorts, in all 422\,400 possible
combinations. However, in this scheme there are many structural zeroes
(impossible combinations) and some random zeros, so the actual number of
combinations we end up with is 24\,326. The data set is thus reduced in
size by 99 per cent.

>>>>

A major theme in demographic and epidemiological studies, is the seemingly persistent effect of social class on mortality. In the present study, we approach this issue by taking a long-term perspective on the development of social inequalities in mortality in the adult and elderly population. Our main focus is on the relative differences. The areas under investigation are the Skellefteå and Umeå regions in the north of Sweden for the period 1801--2013 for Skellefteå and 1901--2013 for Umeå. The main issue is then analysed according to gender and age-group (working age vs retired) and major causes of death (cardiovascular diseases and cancers). Furthermore we put this in the context of how the inequality in mortality is associated with the development of economic inequality in society. The results are discussed in relation to our understanding of the mortality transition and the social determinants of health and mortality, and their implications on some of the most influential hypotheses and concepts in health research.

On the basis of the results we present, we argue that high social class is not necessarily favourable for survival. Mortality risks in different contexts must be understood in the intersection between class and gender. We suggest that health-related behaviour was important not only in present-day societies, but were decisive also in earlier phases of the mortality transition. The results implicate that the association between social class and health is more complex than is assumed in many of the dominant theories in demography and epidemiology.

@@

Another practical problem is the size of that data set:
It contains no less than `r as.integer(round(n.exp + pn.exp))` person years and
 `r n.d + pn.d` deaths.
This makes it very inconvenient to use the nowadays so popular semi-parametric
technique known as
\emph{Cox regression} [@cox72]. Instead we use a parametric model
that allows for reduction of the data
to \emph{sufficient statistics} [@fisher22]. The idea is to keep all
relevant covariates \emph{categorical}, and then create a table with all
possible combinations of levels of the covariates. For each such
combination, the total amount of exposure and the number of deaths are
calculated.  We have two sexes, three marital statuses, ten periods, two
urban/rural groups, four social classes at age 40,  four social classes at age 60,
ten age groups, and 22 cohorts, in all 422\,400 possible
combinations. However, in this scheme there are many structural zeroes
(impossible combinations) and some random zeros, so the actual number of
combinations we end up with is 24\,326. The data set is thus reduced in
size by 99 per cent.
